---
title: "Stream Gage Data Manipulation"
author: "Matt Thacker"
date: "7/1/2020"
output: html_document
---


Lets begin by reading in our stream gage data. The shapefiles referenced are in order: the full GagesII dataset provided by the USGS, a subset of that dataset to the Merrimack basin hydrographic unit as defined by the USGS's watershed boundary dataset, and a subset of the northeast region for the watershed boundary dataset. 

```{r}
#spatial data stuff
library(rgdal)
library(raster)
#data table management
library(dplyr)

#entire gagesII shapefile
gagesII<- shapefile("Gen_Data/gagesII/gagesII_9322_sept30_2011.shp")

#merrimack basin subset
gages.mmack <- shapefile("Gen_Data/Merrimack_Basin/gagesII_mmack_singlepart.shp")

#Watershed boundary dataset
wbd_hu6 <- shapefile("Gen_Data/WBD_01_HU2_Shape/WBDHU6.shp")

#convert watershed boundary dataset from geographic coords. to proj. coords shared with gagesII
wbd_hu6<- spTransform(wbd_hu6, crs(gagesII))

#clip watershed boundary dataset to just merrimack
wbd.mmack <- wbd_hu6[which(wbd_hu6@data$Name == "Merrimack"),]


#subset only gages with complete record from study area
gages.comp <- gages.mmack[which(gages.mmack@data$ACTIVE09 == "yes" & gages.mmack@data$FLYRS1990 == 20),]

#extract attribute table from clipped shapefile
dat.mmack <- gages.comp@data

#subset clipped shapefile to only active stations from 1990-2010
subs.mmack <- dat.mmack[which(dat.mmack$ACTIVE09 == "yes" & dat.mmack$FLYRS1990 == 20),]

#export relevant station ID to textfile to facilate download from USGS data portal, commented out but left in for posterity
#write.table(subs.mmack$STAID, "mmack_ids.txt", sep = ",", row.names = F, quote = F)

#read in station data
usgs <- read.table("Gen_Data/Merrimack_Basin/USGS_Stream_Data/usgs_stream_gages_1990_2010_mmack.txt")

#add correct variable names
colnames(usgs) <- c("agency_cd", "site_no", "parameter_cd", "ts_id", "year_nu", "month_nu", "mean_va")

#filter data a bit
#remove incomplete yer 1990
usgs<- usgs %>% filter(!grepl("1990", year_nu))

#remove the 1990s to allow for better precip data coverage
usgs_2k <- usgs %>% filter(!grepl("199", year_nu))

```




Now lets calculate the variables of interest. To begin with, Ive chosen to look at annual variability within the flow of the stream as that is something likely to be related to the dam removal information I'm interested in exploring. I've tried multiple ways to represent that variability. First, dividing the highest total monthly flow by the lowest total monthly flow in each year is represented by $\delta_{fl}$. Second, the annual variance of the monthly flow totals is represented by $\sigma^2_{fl}$.

```{r}
#work with list outputs
library(rlist)

#generate vector of unique station ids
ids <- unique(usgs$site_no)

#each station has seperate data frame now
stations <- list()
for (i in 1:length(ids)){
  stations[[i]] <- data.frame(usgs[which(usgs$site_no == ids[i]),])
}

#apply solution to the same problen
stationSEP <- function(x){   #function to do what is in the for loop, x is a single value used for indexing within unique IDs list
  out <- data.frame(usgs[which(usgs$site_no == ids[x]),])
  return(out)
}

#vector to apply stationSEP over
itt.station <- 1:length(ids)

#create list of individual station data frames
stations<- lapply(itt.station, stationSEP)


#create final data matrix of annual variability for each gage

#generate vector of unique years
years <- unique(usgs$year_nu)

#vector to apply gage.agg over
itt.agg <- 1:length(years)


#function to extract d.fl values for each gage and year from a given station table, x is a value used for unique year index and n is the index in "stations" of given station table
gage.monthly <- function(x,n){
  #break year from station
  temp.year <-stations[[n]][which(stations[[n]]$year_nu == years[x]), "mean_va"]
  #seasonal calcs
  temp.win <- temp.year[1:3]
  temp.spr <- temp.year[4:6]
  temp.sum <- temp.year[7:9]
  temp.fall <- temp.year[10:12]
  temp.seasons <- c(mean(temp.win), mean(temp.spr), mean(temp.sum), mean(temp.fall))
  dfl <- max(temp.seasons)/min(temp.seasons)
  sig2 <- var(temp.seasons)
  out<- c(dfl,sig2)
  names(out) <- c("DFL.seasonal", "sig2")
  return(out)
}

#apply gage.monthly over each station table, y is index for n in gage.monthlys
rows.monthly<- function (y){
  dfl.vec <- sapply(itt.agg,gage.monthly, n=itt.station[y])
  return(dfl.vec) #vector of dfl for each year at a given gage
}

#create list of calculated vectors for each gage
dat.mmack <- lapply(itt.station, rows.monthly)

##organize outputs
#pull dfl values from list for each gage
dfl.extract <- function(x){
  x[grep("DFL", rownames(x)),]
}

dfl.mmack <-list.rbind(lapply(dat.mmack, dfl.extract))
colnames(dfl.mmack) <- years
rownames(dfl.mmack) <- ids


#pull sig2 vals
sig2.extract <- function(x){
  x[grep("sig2", rownames(x)),]
}
sig2.mmack <-list.rbind(lapply(dat.mmack, sig2.extract))
colnames(sig2.mmack) <- years
rownames(sig2.mmack) <- ids

##write outputs to files for future use, commented out but left for posterity
write.table(dfl.mmack, "Gen_Data/calced_vars/dfl_season_mmack.txt")
write.table(sig2.mmack, "Gen_Data/calced_vars/sig2_season_mmack.txt")

```

